Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-07-11 analysis/paper/paper.qmd
editing

---
title: "Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com, liyingw@asihp.net
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous peoples often had profound impacts on Indigenous societies and cultures in many parts of the world. In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foreign contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century. This study focuses on local subsistence through faunal remains to explore how Indigenous communities responded to trade activities driven by Europeans and Han Chinese, particularly the high demand for natural resources such as deer. By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research examines Indigenous participation in the deerskin trade and their adaptive subsistence strategies. The results suggest cultural persistence in deer consumption, along with increased reliance on suids during the European period as demand for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely explored.

@@

Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percussion scars [@Haynes1983; @Gifford2018; @Fisher1995], were frequently observed on long bones across different layers (@fig-fracture-img: a-d). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. Chewing marks, represented by a series of superficial elongated scores and circular depressions (@fig-fracture-img: e) [@Bello2023], were observed on a few bones, suggesting the evidence of carnivore gnawing. Additionally, natural post-depositional alterations, such as weathering and root etching [@Gifford2018natural], were observed on approximately one-third of the bones across all layers (@fig-fracture-img: c, d; @tbl-deer-modify). These post-depositional impacts may have contributed to an underestimation of cut marks.

@@

<!--deer hides-->
According to documentary sources, deerskin trade flourished in East Asia from the 14th century and was monopolized by the Dutch in the 17th century, when large quantities of hides, particularly from western Taiwan, were exported to Japan [@Cheng2017; @Hollmann1991; @Wang1981]. Archaeological evidence of hide preparation can be supported by cut marks found on the lower limbs of deer at Kiwulan. Moreover, foreign goods found at Kiwulan, such as glass beads, porcelains, and metal objects, might be the result of trade with outsiders [@Wang2020ornament; @Chen2007]. These findings indicate that northeastern Taiwan was also actively engaged in the long-distance trade networks stimulated by European involvement. Although the frequency of lower limbs with skinning marks did not increase substantially during the European period, a disproportionate rise in head parts might imply the use of deer brains for hide tanning in addition to serving as a nutritional resource. Brain tanning is a method for softening and preserving hides used by many Indigenous cultures across North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. In Taiwan, the use of peanuts or pig lard as tanning materials has been culturally revived by the present-day Bunun people in the central highlands based on oral history, with occasional use of animal brains [@Li1999, pp. 48; @Xu1994]. Although the tanning methods once practiced by the Kavalan people remain unclear, brain tanning is a plausible way, given the availability of deer heads from hunting and the large number of cranial fragments found at Kiwulan during the European period. However, further archaeological evidence is still needed to fully support this possibility.

@@

The deerskin trade continued after the Dutch left, but it was much smaller. In the 18th century, about 10,000 pieces were exported annually, far less than the 100,000 pieces exported during the European period [@Nakamura1997; @Tsao2011]. The shrinking trade market and declining population at Kiwulan may have contributed to the reduced reliance of suids, as shown in their lower proportion in the post-European period and eventual absence in the Chinese period. It should be noted that muntjac were consistently part of local subsistence, though in small numbers, until the early Chinese period. Their skins were used by many Indigenous groups such as the Tsou and Bunun for making clothing, similar to the use of deerskins [@Li1999], which may have also been practiced at Kiwulan. Although both suids and muntjacs served as substitutes for deer, their differing frequencies and ubiquity over time suggest that suids were the primary supplementary meat source, while muntjact were less frequently consumed but may have been valued for their hides in daily clothing.

@@

# Conclusion
The zooarchaeological evidence at Kiwulan reveals Indigenous participation in regional trade networks and their adaptation strategies by situating faunal remains within the context of foreign presence, including the Europeans in the 17th century and Chinese in the 19th century. The frequencies, butchery patterns, and modifications on deer bones from Kiwulan indicates an active deer processing that was intensified during the European period. Deer consumption at Kiwulan throughout most of time suggests cultural continuity, while the strategic reliance on suids during the European period reflects adaptation to high trade demands for deer products. On the other hand, the incorporation of muntjac into the diet during the Chinese period likely resulted from the overhunting of sika deer. These findings highlight the adaptation of Kiwulan residents to foreign impacts and environmental changes while maintaining cultural traditions. Through the long-term dynamics of animal use at Kiwulan, this research demonstrates that even Indigenous communities in peripheral areas with weak colonial control were profoundly impacted by colonial activities than previously thought. Future research expanding to other Indigenous settlements in northeastern Taiwan, with additional dietary evidence such as botanical remains and organic residue analysis, could provide deeper insights into regional dietary patterns and human-environment interactions.

>>>>

---
title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com, liyingw@asihp.net
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous peoples often had profound impacts on Indigenous societies and cultures in many parts of the world. In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foreign contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century. This study focuses on local subsistence through faunal remains to explore how Indigenous communities responded to trade activities driven by Europeans and Han Chinese, particularly the high demand for natural resources such as deer. By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research examines Indigenous participation in the deerskin trade and their adaptive subsistence strategies. The results suggest cultural persistence in deer consumption, along with increased reliance on suids during the European period as demand for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely explored.

@@

Burned Bones, normally showing charring marks and color changes, peaked in CL6 at 40% and gradually decreased to less than 10% after CL4 (@fig-cutmark-img: d). Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percussion scars [@Haynes1983; @Gifford2018; @Fisher1995], were commonly observed on long bones (@fig-fracture-img: a-d). These fractures indicate intentional marrow extraction. Specifically, CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. Chewing marks, represented by a series of superficial elongated scores and circular depressions (@fig-fracture-img: e) [@Bello2023], were observed on a few bones, suggesting the evidence of carnivore gnawing. Additionally, natural post-depositional alterations, such as weathering and root etching [@Gifford2018natural], were observed on approximately one-third of the bones across all layers (@fig-fracture-img: c, d; @tbl-deer-modify). These post-depositional impacts may have contributed to an underestimation of cut marks.

@@

<!--deer hides-->
The deerskin trade flourished in East Asia from the 14th century and was monopolized by the Dutch in the 17th century, when large quantities of hides, particularly from western Taiwan, were exported to Japan according to documentary sources [@Cheng2017; @Hollmann1991; @Wang1981]. Skinning marks on deer lower limbs at Kiwulan provide additional archaeological support for the deerskin trade, which aligns with findings from Iron Age contexts in central Taiwan [@Yen2017]. Moreover, foreign goods at Kiwulan, such as beads, porcelains, and metal objects, could be the result of trade with outsiders [@Wang2020ornament; @Chen2007]. These findings indicate that northeastern Taiwan was also actively engaged in the long-distance trade networks stimulated by European involvement. Although the frequency of skinning marks did not increase substantially during the European period, a disproportionate rise in head parts might imply the use of deer brains for hide tanning, in addition to serving as a nutritional resource. Brain tanning is a method for softening and preserving hides used by many Indigenous cultures across North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. In Taiwan, the use of peanuts or pig lard as tanning materials has been culturally revived by the present-day Bunun people in the central highlands based on oral history, with occasional use of animal brains [@Li1999, pp. 48; @Xu1994]. Although the tanning methods once practiced by the Kavalan people remain unclear, brain tanning is a plausible way, given the availability of deer heads from hunting and the large number of cranial fragments found at Kiwulan during the European period. However, further archaeological evidence is still needed to fully support this possibility.

@@

The deerskin trade continued after the Dutch left, but on a much smaller scale. In the 18th century, about 10,000 hides were exported annually, far less than the 100,000 exported during the European period [@Nakamura1997; @Tsao2011]. The shrinking trade market and declining population at Kiwulan may have led to reduced demand for exporting deer products, which subsequently decreased reliance on suids, as reflected in the lower frequencies of suid remains in the post-European period and their eventual absence in the Chinese period. Unlike suids, muntjacs remained a small part of local subsistence until the early Chinese period, possibly due to their value for hide. Many Indigenous groups in Taiwan, including the Tsou and Bunun, used muntjac skins to make clothing and shoes [@Li1999], which may have also been practiced at Kiwulan. Although it is possible that both suids and muntjacs were consumed as supplements for deer, their patterns of use suggest that suids were likely the main source of additional meat, whereas muntjacs might be valued more for their hides, especially as deer populations declined.

@@

# Conclusion
The zooarchaeological evidence at Kiwulan reveals Indigenous participation in large-scale trade networks and their adaptation strategies by situating faunal remains within the context of foreign presence, including the Europeans in the 17th century and Chinese in the 19th century. The frequencies, butchery patterns, and modifications on deer bones from Kiwulan indicate an active deer processing that was intensified during the European period. Deer consumption at Kiwulan throughout most of time suggests cultural continuity, while the strategic reliance on suids during the European period reflects adaptation to high trade demands for deer products. On the other hand, the incorporation of muntjac into the diet during the Chinese period likely resulted from the overhunting of sika deer. Through the long-term dynamics of animal use at Kiwulan, this case study highlights the active responses of its residents to foreign impacts and environmental changes while maintaining cultural traditions. It also demonstrates that Indigenous communities in peripheral areas under limited colonial control were profoundly affected by colonial activities than previously thought. In the future, broader studies incorporating other Indigenous sites in northeastern Taiwan or additional evidence such as botanical remains and residue analysis could offer deeper insights into regional diets and human-environment interactions.
